---
title: "Vignette 2. Psychometric meta-analysis with metaConvert (COSMIN framework)"
author: "Gosling CJ, Cortese S, Solmi M, Haza B, Vieta E, Delorme R, Fusar-Poli P, Radua J"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Vignette 2. Psychometric meta-analysis with metaConvert (COSMIN framework)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{=html}
<style type="text/css">

*{
  font-family: "Gill Sans", sans-serif;
}

h1.title {
  font-weight: 700;
  font-size: 2.2rem;
  padding-top: 0rem;
  margin-top: 0rem;
  border-top: none;
}

#TOC {
  width: 100%;
}

h1{
  font-weight: 550;
  font-size: 1.9rem;
  border-top: 1px solid black;
  margin-top: 3rem;
  padding-top: 2rem;
}


p{
  line-height: 1.4rem;
}
</style>
```

```{r echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(metaConvert)
library(metafor)

show_table <- function(x, ..., digits = 3) {
  if (is.data.frame(x)) {
    num <- vapply(x, is.numeric, logical(1))
    x[num] <- lapply(x[num], round, digits)
  }
  if (requireNamespace("DT", quietly = TRUE)) DT::datatable(x, ...)
  else knitr::kable(x)
}
```

# Introduction

The **COSMIN** (COnsensus-based Standards for the selection of health Measurement INstruments) framework provides a standardised taxonomy of measurement properties for evaluating patient-reported outcome measures (PROMs). Conducting a **meta-analysis of measurement properties** requires computing, transforming, and pooling a variety of effect sizes that are specific to psychometric studies.

The `metaConvert` package provides functions for several COSMIN measurement properties. The table below summarises the coverage:

| COSMIN property | Coverage | metaConvert function(s) |
|:---|:---:|:---|
| Internal consistency | Full | `convert_df(measure = "alpha")` |
| Test-retest reliability | Full | `convert_df(measure = "icc")` |
| Measurement error | Full | `compute_sem()`, `compute_sdc()` |
| Criterion validity | Full | `convert_df(measure = "r")` + `es_disattenuate()` |
| Construct validity | Full | `convert_df(measure = "r")` + `es_disattenuate()` |
| Responsiveness | Full | `reliability_change_score()` + `es_disattenuate()` |
| Floor/ceiling effects | Full | `convert_df(measure = "prop")` |
| Structural validity | Not covered | Requires OSMASEM (metaSEM / OpenMx) |
| Diagnostic accuracy | Not covered | Requires bivariate GLMM (mada / lme4) |
| Cross-cultural validity | Partial | DIF log-odds via OR pipeline; invariance is descriptive |
| Interpretability (MIC) | Partial | Raw MIC via user-provided ES columns |

Throughout this vignette, we use the **`df.psychom`** dataset distributed with metaConvert. It simulates a COSMIN systematic review of a fictional 9-item self-report PROM called the **Mental Health Wellbeing Scale (MHWS)**, scored 0--27, validated across multiple populations and languages.

```{r}
data(df.psychom)
```

```{r, eval=FALSE}
str(df.psychom)
```

```{r, echo=FALSE}
show_table(df.psychom, options = list(
    scrollX = TRUE,
    dom = c('t'),
    scrollY = "300px",
    pageLength = 50,
    ordering = FALSE,
    rownames = FALSE,
    columnDefs = list(
                  list(width = '130px',
                       targets = "_all"),
                  list(className = 'dt-center',
                                     targets = "_all"))))
```

<br>

Each row represents one study reporting one measurement property. The `outcome` column identifies the COSMIN property assessed. Columns not relevant to a study's property are `NA`, reflecting the typical sparsity of real-world data extraction sheets.

# Data preparation

Before data extraction, you can generate a template extraction sheet for each psychometric measure. These templates list all column names that `metaConvert` recognises.

```{r, eval=FALSE}
# Extraction sheet for Cronbach's alpha
data_extraction_sheet(measure = "alpha", extension = "data.frame")

# Extraction sheet for ICC
data_extraction_sheet(measure = "icc", extension = "data.frame")

# Extraction sheet for correlations (Pearson or Spearman)
data_extraction_sheet(measure = "r", extension = "data.frame")

# Extraction sheet for proportions
data_extraction_sheet(measure = "prop", extension = "data.frame")
```

The key psychometric columns are:

| Column | Type | Used by |
|:---|:---|:---|
| `cronbach_alpha` | numeric | `es_from_cronbach_alpha()` |
| `n_items` | numeric | `es_from_cronbach_alpha()` |
| `icc` | numeric | `es_from_icc()` |
| `n_measurements` | numeric | `es_from_icc()` |
| `icc_type` | character | `es_from_icc()` -- `"agreement"` or `"consistency"` |
| `spearman_r` | numeric | `es_from_spearman_rho()` |
| `prop` | numeric | `es_from_prop_single_group()` |
| `n_cases` | numeric | `es_from_prop_single_group_counts()` |


# Internal consistency

Internal consistency is typically assessed using **Cronbach's alpha**. For meta-analysis, the **Bonett (2002) variance-stabilising transformation** is recommended:

$$T(\alpha) = \ln(1 - \alpha), \quad \text{SE} = \sqrt{\frac{2k}{(k-1)(n-2)}}$$

where $k$ is the number of items and $n$ is the sample size (Bonett, 2002). This transformation normalises the sampling distribution and stabilises the variance, which is particularly important when alpha values are close to 1.

## Effect size computation

```{r}
# Subset internal consistency studies
dat_alpha <- df.psychom[df.psychom$outcome == "internal_consistency", ]

# Compute effect sizes with Bonett transformation (default)
res_alpha <- convert_df(dat_alpha, measure = "alpha",
                        verbose = FALSE, split_adjusted = FALSE)
summary_alpha <- summary(res_alpha, flags = TRUE, guidance = TRUE)
```

```{r, eval=FALSE}
summary_alpha[, c("author", "es", "se", "es_ci_lo", "es_ci_up", "info_used")]
```

```{r, echo=FALSE}
show_table(summary_alpha[, c("author", "es", "se", "es_ci_lo", "es_ci_up", "info_used")],
              options = list(scrollX = TRUE, dom = 't', ordering = FALSE,
                             rownames = FALSE, pageLength = 50,
                             columnDefs = list(
                               list(width = '130px', targets = "_all"),
                               list(className = 'dt-center', targets = "_all"))))
```

<br>

The `es` column contains Bonett-transformed values $\ln(1 - \alpha)$, which are always negative (since $\alpha \in [0, 1)$). To **back-transform** for reporting:

$$\hat{\alpha} = 1 - \exp(T)$$

## Meta-analysis

```{r, fig.width=7, fig.height=4}
# Random-effects meta-analysis on Bonett-transformed scale
ma_alpha <- metafor::rma(yi = es, sei = se, data = summary_alpha,
                         method = "REML")

# Forest plot
metafor::forest(ma_alpha, slab = paste0(summary_alpha$author, " (", dat_alpha$year, ")"),
                xlab = "Bonett-transformed alpha: ln(1 - alpha)",
                header = TRUE)
```

```{r}
# Back-transform pooled estimate, CI and prediction interval.
# reliability_backtransform() accepts the rma object directly and performs the
# bound swap for you.
bt <- reliability_backtransform(ma_alpha)
bt

pooled_alpha <- bt$reliability
```

Note that the CI bounds are **swapped** when back-transforming, because $1 - \exp(x)$ is a decreasing function of $x$: the *upper* bound on the transformed scale is the *lower* bound on the alpha scale. Done by hand that is

```{r, eval=FALSE}
1 - exp(ma_alpha$beta[1])   # pooled alpha
1 - exp(ma_alpha$ci.ub)     # LOWER alpha bound  <- from the UPPER T bound
1 - exp(ma_alpha$ci.lb)     # UPPER alpha bound  <- from the LOWER T bound
```

<div style="background:#fff4f4;border-left:4px solid #c0392b;padding:0.6em 1em;margin:1em 0;">
**Do not use `metafor::transf.iabt` on `metaConvert` output.** `metafor`'s `measure = "ABT"` stores $-\ln(1-\alpha)$ where `metaConvert` follows Bonett (2002) and stores $+\ln(1-\alpha)$. The standard errors are identical but every sign is opposite, so `metaConvert`'s values are always negative and `transf.iabt` -- which clamps negative inputs to zero -- maps the whole pool to exactly `0`, with no error, no warning and no `NA`. Use `reliability_backtransform()`.
</div>

The **prediction interval**, not the confidence interval, is what a reliability-generalization review is usually reporting: reliability is a property of scores in a sample, so the question is what alpha to expect in a *new* sample. `reliability_backtransform()` returns `reliability_pi_lo` / `reliability_pi_up` whenever the fitted model carries one.

A caution on heterogeneity: do **not** read $I^2$ here as "these alphas disagree". The Bonett sampling variance $2k/((k-1)(n-2))$ does not involve alpha at all, so $I^2$ is largely a function of the sample sizes -- one fixed set of alphas between .80 and .86 gives $I^2 = 0\%$ at $n = 50$ and $97\%$ at $n = 5000$. Report the back-transformed prediction interval and the observed spread of alphas instead.

The pooled reliability estimate can later serve as input for **disattenuation corrections** in criterion validity, construct validity, and responsiveness analyses (see below).

## Choosing the transformation

`alpha_to_es` offers three scales, and the choice is not cosmetic.

| `alpha_to_es` | stores | sampling variance | notes |
|:---|:---|:---|:---|
| `"bonett"` (default) | $\ln(1-\alpha)$ | $2k/((k-1)(n-2))$ | does **not** involve $\alpha$ |
| `"hakstian_whalen"` | $1-(1-\alpha)^{1/3}$ | $18k(n-1)(1-\alpha)^{2/3}/((k-1)(9n-11)^2)$ | involves $\alpha$ |
| `"raw"` | $\alpha$ | $(1-\alpha)^2 \cdot 2k/((k-1)(n-2))$ | skewed near 1; not recommended |

Because the Bonett variance is free of $\alpha$, two studies with the same $n$ and the same $k$ receive **identical weight** whether their alpha is .85 or .98. The Hakstian--Whalen variance does involve $\alpha$ and separates them --- a 3.83:1 weight ratio at $n = 200$, $k = 10$:

```{r}
wr <- function(m) {
  x <- es_from_cronbach_alpha(c(0.85, 0.98), n_sample = 200, n_items = 10, alpha_to_es = m)
  (x$alpha_se[1] / x$alpha_se[2])^2
}
c(bonett = wr("bonett"), hakstian_whalen = wr("hakstian_whalen"))
```

Hakstian & Whalen (1976) is the transformation Rodriguez & Maeda (2006) recommend and that most published **reliability-generalization** syntheses use, so choose it when you need to reproduce or audit that literature. It is stored in `metafor`'s increasing orientation (identical to `measure = "AHW"` in both `yi` and `vi`), so higher values mean higher reliability and moderator coefficients read in the same direction as alpha. Only the Bonett scale swaps the confidence bounds on back-transformation; `reliability_backtransform()` handles both.

## Heterogeneity, and what not to report

Two cautions specific to reliability pools, neither of which the package can enforce for you.

**$I^2$ is not "how much these alphas disagree".** Under the default Bonett scale the sampling variance $2k/((k-1)(n-2))$ does not involve alpha, so $I^2$ is largely a function of the sample sizes. Holding one fixed set of alphas between .80 and .86 and varying only $n$:

```{r}
a <- c(.80, .82, .83, .84, .85, .86)
sapply(c(50, 200, 1000, 5000), function(n) {
  e <- es_from_cronbach_alpha(a, n_sample = n, n_items = 10)
  round(metafor::rma(yi = e$alpha, sei = e$alpha_se, method = "REML")$I2, 1)
})
```

Same alphas throughout. Report the back-transformed **prediction interval** and the observed spread of alphas instead.

**Do not run funnel plots, trim-and-fill, or Egger's test on this pool.** Those diagnostics assume the effect's expectation depends on the sample size; the expectation of alpha does not (only its precision does), so an asymmetry test is not interpretable here. Edelsbrunner et al. (2025) found trim-and-fill pointing in the *opposite* direction to three valid non-funnel diagnostics on the same data. The reliability-generalization analogue of publication bias is **selective reporting of reliability** --- compare the characteristics of studies that did and did not report alpha at all --- and **reliability induction**, which `summary(flags = TRUE)` screens for automatically.

# McDonald's omega

Omega is increasingly recommended over alpha because it does not assume essential tau-equivalence (equal true-score loadings). That same freedom is why it is entered differently.

**Omega has no closed-form sampling variance in $(n, k)$.** Both alpha variances above descend from the Feldt (1965) / Kristof (1963) result that $(1-r)/(1-\alpha)$ is distributed as $F$, which holds *only* under tau-equivalence. Omega's asymptotic variance depends on the full covariance matrix of the estimated factor loadings and error variances (Raykov, 2002), which primary studies never print, and the recommended interval is a bootstrap one computed from raw data (Kelley & Pornprasertmanit, 2016). So `es_from_omega()` takes the uncertainty **from what the study reported** --- `omega_se` (on the natural scale) or `omega_ci_lo` / `omega_ci_up`:

```{r}
dat_omega <- data.frame(
  study_id    = paste0("S", 1:4),
  omega       = c(0.86, 0.91, 0.78, 0.88),
  omega_se    = c(0.021, NA, 0.035, NA),
  omega_ci_lo = c(NA, 0.87, NA, 0.84),
  omega_ci_up = c(NA, 0.94, NA, 0.92),
  omega_type  = "total",
  n_sample    = c(300, 450, 180, 260),
  n_items     = 12
)
res_omega <- convert_df(dat_omega, measure = "omega", verbose = FALSE,
                        split_adjusted = FALSE)
summary_omega <- summary(res_omega, digits = 15)
ma_omega <- metafor::rma(yi = es, sei = se, data = summary_omega, method = "REML")
reliability_backtransform(ma_omega)
```

A confidence interval is read at its **bounds**, not as a half-width, so an asymmetric bootstrap interval maps correctly (BCa is transformation-respecting). The interval is then summarised by one standard error, from which a symmetric Wald interval is rebuilt: the width is preserved, the asymmetry is not.

A bare omega with neither an SE nor a CI keeps its point estimate and gets `se = NA`, so the study stays visible and countable in the pipeline (and in your reporting-rate accounting) but is not pooled. `summary(guidance = TRUE)` names what is missing.

## Record how omega was estimated

Omega is not one computation, and **which one produced the number matters more than most of the moderators you will code**. Zinbarg et al. (2006) ran several estimators over identical data:

| estimator | bias in $\omega_h$ |
|:---|:---|
| first principal component | **≈ +0.40** |
| hierarchical (bifactor) CFA | ≈ +0.02 |

Revelle and Zinbarg (2009) record the same point from the other side: EFA returns $\omega_h = .04$ on data where CFA returns exactly $0.0$. A bias of that size is larger than any moderator effect a reliability-generalization review is likely to find, so a pool that mixes estimators can manufacture a "language effect" or an "age effect" that is purely an artefact of which software each author ran.

Record it in `omega_estimator`. In practice it is usually codeable from the software a paper reports:

| the paper says | code it as |
|:---|:---|
| `psych::omega()` | `efa_schmid_leiman` (that function defaults to EFA + Schmid–Leiman) |
| a bifactor / hierarchical CFA | `cfa_bifactor` |
| a one-factor CFA, congeneric model | `cfa_1factor` |
| first principal component | `first_pc` |
| principal-axis factoring | `first_pf` |
| `semTools` / `MBESS` / `lavaan`, model not stated | `unspecified` — these are CFA-based, but they do **not** tell you whether the fitted model was unidimensional or bifactor |
| nothing | `unspecified` |

The column is case-insensitive and accepts the obvious synonyms (`bifactor`, `Schmid-Leiman`, `PCA`, …).

```{r}
mixed_est <- data.frame(study_id = c("A", "B"), omega = c(0.86, 0.91), omega_se = 0.02,
                        omega_estimator = c("bifactor", "PCA"),
                        n_sample = 300, n_items = 12)
summary(convert_df(mixed_est, measure = "omega", verbose = FALSE,
                   split_adjusted = FALSE), flags = TRUE)$flags[1]
```

The flag fires only when two **known, different** estimators coexist. Rows left `unspecified` are never counted as mixing — most primary studies do not name their estimator, and treating that as a conflict would make the check fire on nearly every real dataset. A pool that is mostly `unspecified` is a *reporting-quality* observation for your review to make, and worth stating: it means the estimator is a source of heterogeneity you cannot rule out.

## Two extraction traps

Both produce a wrong number silently, and neither can be detected downstream.

**1. Do not recompute omega from a printed loadings table unless the residual variances are printed too.** `lavaan`'s `std.lv = TRUE` fixes the *factor* variance to 1, not the item variances, so the "Estimate" column is **unstandardised**. Assuming $\theta_j = 1 - \lambda_j^2$ from such a table gives 0.7725 for the worked example in Flora (2020) whose published omega is **0.5999** — an error of +0.17. If you need to recompute, you need $\sum\lambda_j$ *and* $\sum\theta_j$, both printed.

**2. On a bifactor model, the output row labelled `omega` is not $\omega_h$.** In `semTools` it is the total — 0.97 in Flora's example, where $\omega_h$ is 0.91. Copy the row you actually mean, and record which one in `omega_type`.

## Two things not to pool

**`omega_type` is an estimand, not a label.** `"total"` is the proportion of variance due to all common factors; `"hierarchical"` is the proportion due to the *general* factor alone, and is systematically smaller. A pool mixing them is flagged:

```{r}
mixed <- data.frame(study_id = c("A", "B"), omega = c(0.86, 0.72), omega_se = 0.02,
                    omega_type = c("total", "hierarchical"), n_sample = 300, n_items = 12)
summary(convert_df(mixed, measure = "omega", verbose = FALSE, split_adjusted = FALSE),
        flags = TRUE)$flags
```

**Do not pool omega with alpha.** Under a congeneric model $\omega_{total} \ge \alpha$, with equality only under exact tau-equivalence, so they are different quantities. There is a second reason specific to this package: alpha's standard errors are analytic functions of $(n, k)$ while omega's come from reported intervals, so the *weighting scheme* differs systematically by metric --- metric would be perfectly confounded with precision. Run two pools and report them side by side. If some studies report both, estimate the gap as a **within-study paired contrast** rather than by differencing two separately pooled means: the two coefficients are computed on the same data and are near-perfectly dependent, so treating the pools as independent badly understates the standard error of the difference.

## Alternative: raw alpha

An alternative is to meta-analyse raw Cronbach's alpha values directly (`alpha_to_es = "raw"`), using the SE formula from Feldt et al. (1987). This is simpler to interpret but the sampling distribution may be skewed when alpha is close to 1. The Bonett transformation is generally recommended.

```{r, eval=FALSE}
# Raw alpha (not recommended for meta-analysis)
res_alpha_raw <- convert_df(dat_alpha, measure = "alpha", alpha_to_es = "raw",
                            verbose = FALSE, split_adjusted = FALSE)
summary(res_alpha_raw)
```


# Test-retest reliability

Test-retest reliability is assessed using the **intraclass correlation coefficient (ICC)**. The same **Bonett (2002) transformation** is applied:

$$T(\text{ICC}) = \ln(1 - \text{ICC})$$

The SE of the **transformed** ICC is obtained by the delta method: the $(1-\rho)^2$ term from the raw ICC variance cancels with the $1/(1-\rho)^2$ from the transformation derivative. For a single-measure ICC, this leading-order variance is the **same** for the two-way absolute-agreement ICC(2,1) and the two-way consistency ICC(3,1):

- **ICC(2,1) -- absolute agreement** (`icc_type = "agreement"`) and **ICC(3,1) -- consistency** (`icc_type = "consistency"`):
$$\text{SE}(T) = \sqrt{\frac{2(1+(k-1)\rho)^2}{k(k-1)(n-1)}}$$

where $k$ is the number of measurement occasions and $\rho$ is the ICC value. Deriving the ICC(3,1) variance directly from $F_0 = \text{MSR}/\text{MSE}$ (degrees of freedom $n-1$ and $(n-1)(k-1)$) yields $\left[(1+(k-1)\rho)/k\right]^2 \cdot 2k/((k-1)(n-1))$, which reduces to the same expression --- so, contrary to a common simplification, the ICC(3,1) SE is **not** independent of $\rho$ (a Monte Carlo check confirms the $\rho$-dependent form; a $\rho$-free $\sqrt{2/((k-1)(n-1))}$ under-states the SE by up to ~2--3$\times$ in variance for high reliabilities).

**Technical note on the ICC(2,1) formula**: The SE formula labelled "ICC(2,1)" above is technically the analytical variance for the **one-way random ICC(1,1)** model (Shrout & Fleiss, 1979), as derived by Bonett (2002). The exact two-way random ICC(2,1) variance formula involves separate between-rater and residual variance components that are typically unavailable without the raw ANOVA mean squares. Using the one-way formula as a proxy for ICC(2,1) is standard practice in psychometric meta-analysis (see e.g., Bonett, 2002; Donner & Eliasziw, 1987) and provides a reasonable approximation.

## Record the retest interval

A test-retest ICC is not a property of the instrument alone: it depends on **how long
the gap was**. Reliability decays with the interval, and the primary moderator in
published test-retest syntheses is usually the interval itself (Gnambs, 2014; Scharfen
et al., 2018). Polit (2014) treats reporting it as a core recommendation. A pool that
mixes one-week and one-year retests and reports a single number is answering a question
nobody asked.

`metaConvert` does not need a registered column for this, and deliberately does not have
one: any extra column you add to the input survives `convert_df()` unchanged and can be
used as a moderator directly. That includes character values, so `"1 week"` stays
`"1 week"` --- registering the column as numeric would silently coerce it to `NA`.

```{r, eval = FALSE}
# retest_interval_days is your own column: nothing in the package reads it, and
# it passes through untouched.
dat_icc$retest_interval_days <- c(7, 14, 30, 30, 90, 180, 365)

ma_int <- metafor::rma(yi = es, sei = se, mods = ~ log(retest_interval_days),
                       data = summary_icc, method = 'REML')
```

Prefer `log(interval)` to the raw interval: decay is closer to linear on the log scale,
and retest gaps in a real corpus span orders of magnitude.

## Effect size computation

```{r}
# Subset test-retest studies
dat_icc <- df.psychom[df.psychom$outcome == "test_retest", ]

# Compute effect sizes with Bonett transformation (default)
res_icc <- convert_df(dat_icc, measure = "icc",
                      verbose = FALSE, split_adjusted = FALSE)
summary_icc <- summary(res_icc, flags = TRUE)
```

```{r, eval=FALSE}
summary_icc[, c("author", "es", "se", "es_ci_lo", "es_ci_up",
                 "icc_type", "info_used")]
```

```{r, echo=FALSE}
show_table(summary_icc[, c("author", "es", "se", "es_ci_lo", "es_ci_up",
                                "icc_type", "info_used")],
              options = list(scrollX = TRUE, dom = 't', ordering = FALSE,
                             rownames = FALSE, pageLength = 50,
                             columnDefs = list(
                               list(width = '130px', targets = "_all"),
                               list(className = 'dt-center', targets = "_all"))))
```

<br>

**Do not pool agreement and consistency ICCs without deciding to.** ICC(2,1) and ICC(3,1) are different estimands: absolute agreement charges systematic rater differences against the reliability, consistency does not, so a consistency ICC is the larger of the two on the same data and their average estimates neither. The arithmetic is *identical* for both --- `icc_type` has no effect on the estimate or its standard error --- so no numeric diagnostic can reveal the mix. `summary(flags = TRUE)` raises **V43** on every row of a mixed pool for exactly that reason. Split the analysis by `icc_type`, or enter it as a **moderator** and report the contrast.

A second flag applies to the agreement rows themselves. Both types share one leading-order SE formula, which is exact for consistency ICC(3,1) but for agreement ICC(2,1) is a one-way approximation assuming negligible between-rater variance. When raters differ systematically it is anti-conservative, and increasingly so at larger *n*: measured 95% coverage is 0.82 at *n* = 20, 0.67 at 50, 0.32 at 200 and 0.14 at 1000, because ICC(2,1) inherits the between-rater mean square (*k* - 1 df) while the reported SE shrinks like $1/\sqrt{n}$. A large agreement study can therefore carry ~50x too much weight. Where the study reports its own `icc_se` or confidence interval, supply it --- `es_from_icc()` prefers it to the computed value. Where it does not, `convert_df(icc_agreement_se = "drop")` withholds the approximation so the row is left out of the pool rather than over-weighted (see `?es_from_icc`):

```{r}
summary_icc[nzchar(summary_icc$flags), c("author", "icc_type")]
unique(summary_icc$flags[nzchar(summary_icc$flags)])
```

When pooling, `icc_type` can be examined as a **moderator** variable.

## Meta-analysis

```{r, fig.width=7, fig.height=4}
# Random-effects meta-analysis
ma_icc <- metafor::rma(yi = es, sei = se, data = summary_icc,
                       method = "REML")

# Forest plot
metafor::forest(ma_icc, slab = paste0(summary_icc$author, " (", dat_icc$year, ")"),
                xlab = "Bonett-transformed ICC: ln(1 - ICC)",
                header = TRUE)
```

```{r}
# Back-transform pooled estimate
# reliability_backtransform() performs the bound swap for you: 1 - exp(T) is
# monotone decreasing, so the transformed-scale upper bound is the ICC-scale
# lower bound. This step is easy to get wrong by hand, and the mistake is hard to
# spot, because the interval still looks like an interval either way.
bt_icc <- reliability_backtransform(ma_icc)

cat(sprintf("Pooled ICC = %.3f [%.3f, %.3f]\n",
            bt_icc$reliability, bt_icc$reliability_ci_lo, bt_icc$reliability_ci_up))
```

The same two cautions as for alpha apply here, and for the same reason. **Do not read $I^2$ as "these ICCs disagree":** the Bonett ICC sampling variance $2(1+(k-1)\rho)^2/(k(k-1)(n-1))$ is dominated by $n$ and $k$, so $I^2$ largely tracks the sample sizes rather than any disagreement between the ICCs. Report the back-transformed **prediction interval** (`reliability_pi_lo` / `reliability_pi_up`) and the observed spread of ICCs instead. **And do not run funnel plots, trim-and-fill, or Egger's test on this pool:** those diagnostics assume the effect's expectation depends on sample size, and the expectation of an ICC does not --- only its precision does.

## Several ICCs from one study

A study reporting several subscales, or several retest gaps, contributes several
non-independent rows. `aggregate_df()` combines them into one effect size per study
before pooling.

**It returns four columns and drops everything else.** The aggregating factor, the
aggregated effect size and its standard error --- and none of your moderators. A
40-column `summary()` frame comes back with four. This is deliberate (there is no
defined value for a column that varies within a cluster, so the function refuses to
invent one), but it is silent, so it is worth knowing before you wonder where
`year` went.

```{r, eval = FALSE}
# name the moderators you want kept: col_fact for categorical (must be constant
# within the cluster), col_mean and friends for numeric
agg <- aggregate_df(summary_icc, dependence = 'outcomes', agg_fact = 'study_id',
                    es = 'es', se = 'se',
                    col_fact = c('instrument', 'population'),
                    col_mean = 'retest_interval_days')

# or merge them back afterwards, which is simpler for study-level variables that
# cannot vary within a study anyway
agg <- merge(agg, unique(summary_icc[, c('study_id', 'year', 'country')]),
             by = 'study_id')
```

The pooled ICC will feed into the **measurement error** derivation (next section) and **disattenuation** corrections for validity and responsiveness.

Studies that report only Pearson correlations for test-retest reliability (rather than ICC) can be pooled separately using `convert_df(..., measure = "r")`.


# Measurement error

Measurement error is characterised by the **standard error of measurement (SEM)** and the **smallest detectable change (SDC)**. These are not traditional effect sizes but are essential COSMIN measurement properties.

The SEM is derived from matched ICC and SD data:

$$\text{SEM} = \text{SD} \times \sqrt{1 - \text{ICC}}$$

The SDC is derived from the SEM:

$$\text{SDC} = 1.96 \times \sqrt{2} \times \text{SEM}$$

These are computed using **standalone utility functions** (not `convert_df()`), because they require combining reliability and variability from the same sample.

## Study-specific SEM and SDC

```{r}
# Use test-retest studies that report both ICC and SD
dat_me <- df.psychom[df.psychom$outcome == "test_retest" &
                     !is.na(df.psychom$sd_scores), ]

# Compute SEM for each study (delta-method SE included)
sem_results <- compute_sem(
  sd       = dat_me$sd_scores,
  icc      = dat_me$icc,
  n_sample = dat_me$n_sample
)

# Chain to SDC
sdc_results <- compute_sdc(
  sem    = sem_results$sem,
  sem_se = sem_results$sem_se
)
```

```{r, eval=FALSE}
# Combined measurement error table
measurement_error <- data.frame(
  study  = dat_me$study_id,
  n      = dat_me$n_sample,
  sd     = dat_me$sd_scores,
  icc    = dat_me$icc,
  sem    = round(sem_results$sem, 3),
  sem_se = round(sem_results$sem_se, 3),
  sdc    = round(sdc_results$sdc, 3),
  sdc_se = round(sdc_results$sdc_se, 3)
)
measurement_error
```

```{r, echo=FALSE}
measurement_error <- data.frame(
  study  = dat_me$study_id,
  n      = dat_me$n_sample,
  sd     = dat_me$sd_scores,
  icc    = dat_me$icc,
  sem    = round(sem_results$sem, 3),
  sem_se = round(sem_results$sem_se, 3),
  sdc    = round(sdc_results$sdc, 3),
  sdc_se = round(sdc_results$sdc_se, 3)
)
show_table(measurement_error,
              options = list(scrollX = TRUE, dom = 't', ordering = FALSE,
                             rownames = FALSE, pageLength = 50,
                             columnDefs = list(
                               list(width = '100px', targets = "_all"),
                               list(className = 'dt-center', targets = "_all"))))
```

## Pooling SEM

The study-specific SEM values can be pooled using inverse-variance random-effects meta-analysis, with the delta-method SE providing the sampling variance:

```{r, fig.width=7, fig.height=4}
ma_sem <- metafor::rma(yi = sem_results$sem, sei = sem_results$sem_se,
                       method = "REML")

metafor::forest(ma_sem,
                slab = paste0(dat_me$author, " (", dat_me$year, ")"),
                xlab = "Standard Error of Measurement (SEM)",
                header = TRUE)
```

```{r}
cat(sprintf("Pooled SEM = %.3f [%.3f, %.3f]\n",
            ma_sem$beta[1], ma_sem$ci.lb, ma_sem$ci.ub))

# Derive pooled SDC from pooled SEM
pooled_sdc <- 1.96 * sqrt(2) * ma_sem$beta[1]
cat(sprintf("Pooled SDC = %.3f\n", pooled_sdc))
```

For interpretability, the ratio of pooled SDC to pooled MIC (minimal important change) indicates whether the instrument can reliably detect changes that patients consider meaningful. A ratio $< 1$ (SDC < MIC) is desirable.

**Important**: SEM and SDC are expressed in the **raw units of the measurement instrument**. Pooling is only valid when all primary studies use the exact same version of the instrument with the same scoring range. If some studies use the original 0--27 MHWS and others use a modified 0--100 version, their SEMs are on different scales and cannot be pooled directly. In such cases, consider standardising by the score SD (computing the coefficient of variation SEM/SD) or pooling only within homogeneous subgroups.


# Criterion and construct validity

Criterion and construct validity are assessed by pooling **correlations** between the MHWS and comparator instruments. Some studies report Pearson's $r$, others report Spearman's $\rho$. The `metaConvert` pipeline handles both: Spearman correlations are automatically converted to Pearson-equivalent values using the **Rupinski & Dunlap (1996)** formula:

$$r_{\text{Pearson}} = 2 \sin\!\left(\frac{\pi}{6} \times r_{\text{Spearman}}\right)$$

The SE is derived via the delta method, correctly propagating the uncertainty from the conversion.

Results should be reported three ways:

- **(a) Uncorrected**: observed correlations pooled directly
- **(b) Individually corrected**: study-specific disattenuation, then pooled
- **(c) Artifact-distribution corrected**: using the distribution of reliabilities across studies (Hunter & Schmidt method; see the `psychmeta` package)

This vignette demonstrates approaches (a) and (b).

## Effect size computation

```{r}
# Subset criterion and construct validity studies
dat_validity <- df.psychom[df.psychom$outcome %in%
                           c("criterion_validity", "construct_validity"), ]

# convert_df automatically handles both Pearson r and Spearman rho
res_r <- convert_df(dat_validity, measure = "r",
                    verbose = FALSE, split_adjusted = FALSE)
summary_r <- summary(res_r, flags = TRUE)
```

```{r, eval=FALSE}
summary_r[, c("author", "es", "se", "info_used")]
```

```{r, echo=FALSE}
show_table(summary_r[, c("author", "es", "se", "es_ci_lo", "es_ci_up", "info_used")],
              options = list(scrollX = TRUE, dom = 't', ordering = FALSE,
                             rownames = FALSE, pageLength = 50,
                             columnDefs = list(
                               list(width = '130px', targets = "_all"),
                               list(className = 'dt-center', targets = "_all"))))
```

<br>

The `info_used` column distinguishes between `"pearson_r"` and `"spearman_r"` sources, providing traceability for each study's conversion pathway.

## (a) Pooling uncorrected correlations

Meta-analysis of correlations is typically performed on the **Fisher's z** scale, then back-transformed. Here, `convert_df` with `measure = "r"` returns correlations in the `es` column. The SE on the Fisher's z scale is derived via the delta method: $\text{SE}(z) = \text{SE}(r) / (1 - r^2)$, which correctly propagates the uncertainty from any upstream conversion (e.g., Spearman to Pearson):

```{r, fig.width=7, fig.height=5}
# Pool observed (uncorrected) correlations on Fisher's z scale
z_obs <- atanh(summary_r$es)
r_bounded <- pmin(pmax(summary_r$es, -0.9999), 0.9999)
z_se  <- summary_r$se / (1 - r_bounded^2)

ma_r_obs <- metafor::rma(yi = z_obs, sei = z_se, method = "REML")

# Forest plot on correlation scale
metafor::forest(ma_r_obs,
                slab = paste0(summary_r$author, " (", dat_validity$year, ")"),
                xlab = "Fisher's z (observed correlation)",
                header = TRUE)
```

```{r}
# Back-transform to correlation scale
pooled_r_obs <- tanh(ma_r_obs$beta[1])
pooled_r_lo  <- tanh(ma_r_obs$ci.lb)
pooled_r_up  <- tanh(ma_r_obs$ci.ub)

cat(sprintf("(a) Pooled uncorrected r = %.3f [%.3f, %.3f]\n",
            pooled_r_obs, pooled_r_lo, pooled_r_up))
```

## (b) Individual disattenuation then pooling

The `es_disattenuate()` function corrects each observed correlation for measurement error in both the target PROM and the comparator instrument:

$$r_{\text{corrected}} = \frac{r_{\text{observed}}}{\sqrt{\text{rel}_X \times \text{rel}_Y}}$$

```{r}
# Disattenuate using study-specific reliabilities
corrected <- es_disattenuate(
  r             = summary_r$es,
  r_se          = summary_r$se,
  reliability_x = dat_validity$reliability_target,
  reliability_y = dat_validity$reliability_comparator,
  n_sample      = dat_validity$n_sample
)
```

```{r, eval=FALSE}
# Comparison table
data.frame(
  study              = dat_validity$study_id,
  info_used          = summary_r$info_used,
  r_observed         = round(summary_r$es, 3),
  r_corrected        = round(corrected$r_corrected, 3),
  attenuation_factor = round(corrected$attenuation_factor, 3)
)
```

```{r, echo=FALSE}
comp <- data.frame(
  study              = dat_validity$study_id,
  info_used          = summary_r$info_used,
  r_observed         = round(summary_r$es, 3),
  r_corrected        = round(corrected$r_corrected, 3),
  attenuation_factor = round(corrected$attenuation_factor, 3)
)
show_table(comp,
              options = list(scrollX = TRUE, dom = 't', ordering = FALSE,
                             rownames = FALSE, pageLength = 50,
                             columnDefs = list(
                               list(width = '130px', targets = "_all"),
                               list(className = 'dt-center', targets = "_all"))))
```

<br>

The corrected correlations are systematically larger in magnitude, reflecting the removal of attenuation due to measurement error.

**Note on bounds**: Because sampling error exists in both the observed correlation and the reliability estimates, `r_corrected` can occasionally exceed 1.0 in absolute value. When this happens, `es_disattenuate()` returns the corrected correlation as computed (it is not capped), emits a warning, and sets the corrected-r confidence interval and all Fisher's z outputs (`z_corrected`, `z_corrected_se`, and the z confidence bounds) to `NA`: beyond $|r_c| > 0.999$ a bounded $\text{atanh}$ would return the same clamp artefact whatever the input. Such rows are omitted by `metafor::rma()`, so the reliability inputs should be checked before pooling.

However, a corrected correlation exceeding 1.0 is a **red flag** that warrants investigation. Common causes include: (1) the reliability coefficient was estimated from a different population than the study sample; (2) the reliability was based on a very small sample and is itself imprecise; or (3) range restriction artefacts interact with the correction. We recommend running a **sensitivity analysis** excluding studies where $|r_c| > 1$ and comparing results. If multiple studies produce impossible values, an artifact-distribution method (Hunter & Schmidt, 2004; see the `psychmeta` package) may be more appropriate than individual study-level correction, as it uses the *distribution* of reliability estimates rather than point values.

```{r, fig.width=7, fig.height=5}
# Pool corrected correlations on Fisher's z scale
ma_r_corr <- metafor::rma(yi = corrected$z_corrected,
                           sei = corrected$z_corrected_se,
                           method = "REML")

metafor::forest(ma_r_corr,
                slab = paste0(summary_r$author, " (", dat_validity$year, ")"),
                xlab = "Fisher's z (disattenuated correlation)",
                header = TRUE)
```

```{r}
pooled_r_corr <- tanh(ma_r_corr$beta[1])
pooled_r_corr_lo <- tanh(ma_r_corr$ci.lb)
pooled_r_corr_up <- tanh(ma_r_corr$ci.ub)

cat(sprintf("(b) Pooled disattenuated r = %.3f [%.3f, %.3f]\n",
            pooled_r_corr, pooled_r_corr_lo, pooled_r_corr_up))
```

## Using pooled reliability when study-specific data are unavailable

When study-specific reliability data are unavailable, the pooled reliability estimates from the *Internal consistency* and *Test-retest reliability* meta-analyses above can be used:

```{r, eval=FALSE}
# Use pooled alpha and ICC from earlier sections as proxies
es_disattenuate(
  r = 0.65,
  r_se = 0.08,
  reliability_x = pooled_alpha,  # from the Internal consistency section
  reliability_y = 0.85,          # from external published data
  n_sample = 150
)
```


# Responsiveness

Responsiveness is assessed by correlating **change scores** on the MHWS with change scores on a comparator instrument (e.g., CGI-Change, PGIC). Because change scores have lower reliability than single-occasion scores, **disattenuation requires change-score reliability**, not single-occasion reliability.

The **Lord (1963)** formula derives change-score reliability from single-occasion reliability and the pre-post correlation:

$$\text{rel}_{\text{change}} = \frac{\text{rel} - r_{\text{pre-post}}}{1 - r_{\text{pre-post}}}$$

## Effect size computation with change-score correction

```{r}
# Subset responsiveness studies
dat_resp <- df.psychom[df.psychom$outcome == "responsiveness", ]

# Step 1: Compute change-score reliability
change_rel <- reliability_change_score(
  reliability = dat_resp$single_occasion_reliability,
  r_pre_post  = dat_resp$r_pre_post
)

# Step 2: Get responsiveness correlations via convert_df
res_resp <- convert_df(dat_resp, measure = "r",
                       verbose = FALSE, split_adjusted = FALSE)
summary_resp <- summary(res_resp)

# Step 3: disattenuate using the change-score reliability
corrected_resp <- es_disattenuate(
  r             = summary_resp$es,
  r_se          = summary_resp$se,
  reliability_x = change_rel$rel_change,
  reliability_y = dat_resp$reliability_comparator,
  n_sample      = dat_resp$n_sample
)
```

```{r, eval=FALSE}
data.frame(
  study        = dat_resp$study_id,
  r_observed   = round(summary_resp$es, 3),
  rel_single   = dat_resp$single_occasion_reliability,
  r_pre_post   = dat_resp$r_pre_post,
  rel_change   = round(change_rel$rel_change, 3),
  r_corrected  = round(corrected_resp$r_corrected, 3)
)
```

```{r, echo=FALSE}
resp_table <- data.frame(
  study        = dat_resp$study_id,
  r_observed   = round(summary_resp$es, 3),
  rel_single   = dat_resp$single_occasion_reliability,
  r_pre_post   = dat_resp$r_pre_post,
  rel_change   = round(change_rel$rel_change, 3),
  r_corrected  = round(corrected_resp$r_corrected, 3)
)
show_table(resp_table,
              options = list(scrollX = TRUE, dom = 't', ordering = FALSE,
                             rownames = FALSE, pageLength = 50,
                             columnDefs = list(
                               list(width = '110px', targets = "_all"),
                               list(className = 'dt-center', targets = "_all"))))
```

<br>

Note that change-score reliability (`rel_change`) is always lower than single-occasion reliability (`rel_single`), which leads to a **larger disattenuation correction**.

## Meta-analysis

```{r, fig.width=7, fig.height=4}
# Pool observed change-score correlations
z_resp     <- atanh(summary_resp$es)
r_resp_bounded <- pmin(pmax(summary_resp$es, -0.9999), 0.9999)
z_resp_se  <- summary_resp$se / (1 - r_resp_bounded^2)

ma_resp_obs <- metafor::rma(yi = z_resp, sei = z_resp_se, method = "REML")

cat(sprintf("Pooled observed r = %.3f [%.3f, %.3f]\n",
            tanh(ma_resp_obs$beta[1]),
            tanh(ma_resp_obs$ci.lb),
            tanh(ma_resp_obs$ci.ub)))

# Pool disattenuated change-score correlations
ma_resp_corr <- metafor::rma(yi = corrected_resp$z_corrected,
                              sei = corrected_resp$z_corrected_se,
                              method = "REML")

cat(sprintf("Pooled disattenuated r = %.3f [%.3f, %.3f]\n",
            tanh(ma_resp_corr$beta[1]),
            tanh(ma_resp_corr$ci.lb),
            tanh(ma_resp_corr$ci.ub)))
```

## Sensitivity analysis for unknown pre-post correlations

When pre-post correlations are not reported, a sensitivity analysis using plausible values illustrates the impact on the corrected estimates:

```{r}
# Example: one study with r_observed = 0.50, rel_single = 0.85
r_pre_post_grid <- c(0.20, 0.40, 0.60, 0.80)

sensitivity <- data.frame(
  r_pre_post = r_pre_post_grid,
  rel_change = sapply(r_pre_post_grid, function(rpp) {
    reliability_change_score(reliability = 0.85, r_pre_post = rpp)$rel_change
  }),
  r_corrected = sapply(r_pre_post_grid, function(rpp) {
    rc <- reliability_change_score(reliability = 0.85, r_pre_post = rpp)$rel_change
    es_disattenuate(r = 0.50, r_se = 0.08, reliability_x = rc,
                    reliability_y = 0.80, n_sample = 100)$r_corrected
  })
)
sensitivity
```

As the pre-post correlation increases, change-score reliability decreases, and the disattenuation correction becomes larger. At $r_{\text{pre-post}} = 0.80$ the change-score reliability falls to 0.25 and the corrected correlation exceeds 1, which is why `es_disattenuate()` warns and sets the Fisher's z outputs of that row to `NA` (see the note on bounds above). This highlights why reporting pre-post correlations in primary studies is essential.


# Floor and ceiling effects

Floor and ceiling effects can be assessed by meta-analysing the **proportion** of respondents scoring at the minimum or maximum of the scale. Three transformations are available:

- **`"raw"`** (default): raw proportion, SE = $\sqrt{p(1-p)/n}$. CIs bounded to [0, 1].
- **`"logit"`**: logit transform $\log(p/(1-p))$, SE = $\sqrt{1/(np(1-p))}$
- **`"freeman_tukey"`**: Freeman-Tukey double arcsine. Handles zero proportions naturally and stabilises the variance.

## Effect size computation and pooling

```{r}
# Subset floor/ceiling studies
dat_prop <- df.psychom[df.psychom$outcome == "floor_ceiling", ]

# Freeman-Tukey transformation
res_prop <- convert_df(dat_prop, measure = "prop", prop_to_es = "freeman_tukey",
                       verbose = FALSE, split_adjusted = FALSE)
summary_prop <- summary(res_prop)
```

```{r, fig.width=7, fig.height=3.5}
ma_prop <- metafor::rma(yi = es, sei = se, data = summary_prop,
                        method = "REML")

metafor::forest(ma_prop,
                slab = paste0(summary_prop$author, " (", dat_prop$year, ")"),
                xlab = "Freeman-Tukey transformed proportion",
                header = TRUE)
```

```{r}
# Approximate back-transformation (for the pooled estimate)
pooled_prop <- sin(ma_prop$beta[1])^2
cat(sprintf("Pooled proportion (approx.) = %.3f\n", pooled_prop))
```

**Note on the Freeman-Tukey back-transformation**: Of the three transformations, the double arcsine handles boundary proportions best within `metaConvert`, but the back-transformation of the pooled estimate is only approximate. The `sin(x)^2` formula used above ignores the sample sizes, and Schwarzer et al. (2019, *Research Synthesis Methods*) have shown that the Freeman-Tukey back-transformation can distort pooled results when sample sizes are heterogeneous across studies. For proportions near 0 or 1 (as is typical for floor/ceiling effects), a generalised linear mixed model (GLMM) with logit link --- available via `metafor::rma.glmm()` --- may be preferable as it avoids the transformation entirely and directly models the binomial counts.


# Beyond metaConvert

Several COSMIN measurement properties require specialised methods not covered by `metaConvert`:

**Structural validity** requires meta-analytic structural equation modelling (OSMASEM; Jak & Cheung, 2020) to pool inter-item correlation matrices and fit confirmatory factor models. See the `metaSEM` and `OpenMx` packages.

**Diagnostic and screening accuracy** requires the bivariate generalised linear mixed model (Chu & Cole, 2006) for jointly modelling sensitivity and specificity. See the `mada` and `lme4` packages, and the `HSROC` model for summary ROC curves.

**Cross-cultural validity and measurement invariance** is primarily a descriptive synthesis (tabulating configural, metric, scalar invariance levels). When DIF statistics (log-odds ratios) are available, they can be pooled using metaConvert's existing OR pipeline via `convert_df(measure = "or")`.

**Interpretability (MIC)** can be partially handled: if you have MIC estimates and their SEs, enter them as user-provided effect sizes via the `user_es_original_measure_crude`, `user_es_crude`, and `user_se_crude` columns, then use `convert_df()` to organise them alongside other effect sizes.

**Feasibility and acceptability** are synthesised narratively, without quantitative pooling.


# Summary

| COSMIN property | metaConvert step | Key function(s) |
|:---|:---|:---|
| Internal consistency | `convert_df(measure = "alpha")` | `es_from_cronbach_alpha()` |
| Test-retest reliability | `convert_df(measure = "icc")` | `es_from_icc()` |
| Measurement error | Standalone | `compute_sem()` + `compute_sdc()` |
| Criterion / construct validity | `convert_df(measure = "r")` then `es_disattenuate()` | `es_from_pearson_r()`, `es_from_spearman_rho()`, `es_disattenuate()` |
| Responsiveness | `reliability_change_score()` then `es_disattenuate()` | Full chain |
| Floor/ceiling effects | `convert_df(measure = "prop")` | `es_from_prop_single_group()` |

---
