---
title: "Vignette 1. General guidance about metaConvert"
author: "Gosling CJ, Cortese S, Solmi M, Haza B, Vieta E, Delorme R, Fusar-Poli P, Radua J"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Vignette 1. General guidance about metaConvert}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{=html}
<style type="text/css">

*{
  font-family: "Gill Sans", sans-serif;
}

h1.title {
  font-weight: 700;
  font-size: 2.2rem;
  padding-top: 0rem;
  margin-top: 0rem;
  border-top: none;
}

#TOC {
  width: 100%;
}

h1{
  font-weight: 550;
  font-size: 1.9rem;
  border-top: 1px solid black;
  margin-top: 3rem;
  padding-top: 2rem;
}


p{
  line-height: 1.4rem;
}
</style>
```

```{r echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(metaConvert)

# DT is a Suggests-only dependency, so it is absent on the check flavour that
# installs hard dependencies only. Fall back to a static table there.
show_table <- function(x, ..., digits = 3) {
  # summary() returns es/se/CI at full precision on purpose: se becomes an
  # inverse-variance weight downstream, so rounding it in the data would bias
  # every pooled estimate. Rounding is therefore a display-time job, done here.
  if (is.data.frame(x)) {
    num <- vapply(x, is.numeric, logical(1))
    x[num] <- lapply(x[num], round, digits)
  }
  if (requireNamespace("DT", quietly = TRUE)) DT::datatable(x, ...)
  else knitr::kable(x)
}
```


# Step 1. Protocol stage

If you have not yet registered your protocol, you can benefit from our tools to select **a priori** the type of input data that could be extracted to estimate an effect size.

- start by determining the effect size measure (SMD, OR, RR, etc.) you plan to estimate
- then, identify all types of input data that can be used to estimate this effect size measure
```{r, eval = FALSE}
see_input_data(measure = "or")
```

```{r, echo=FALSE}
show_table(see_input_data(measure = "or", type_of_measure = "natural", extension="data.frame"), options = list(  
    scrollX = TRUE,
    dom = c('t'),
    scrollY = "300px", 
    pageLength = 500,
    ordering = FALSE,
    rownames = FALSE,
    columnDefs = list(
                  list(width = '130px',
                       targets = "_all"),
                  list(className = 'dt-center', 
                                     targets = "_all"))))
```

<br><br>

- last, generate a data extraction sheet for this effect size measure
```{r, eval = FALSE}
data_extraction_sheet(measure = "or")
```

```{r, echo=FALSE}
show_table(data_extraction_sheet(measure = "or", extension="data.frame"), options = list(  
    scrollX = TRUE,
    dom = c('t'),
   ordering = FALSE,
    rownames = FALSE,
    columnDefs = list(
                  list(width = '130px',
                       targets = "_all"),
                  list(className = 'dt-center', 
                                     targets = "_all"))))
```


# Step 2. Dataset comparison

When data extraction has been performed in duplicate, our tools offer the possibility to flag the differences between the two datasets. For this example, we will use two datasets (df.compare1 and df.compare2) distributed with metaConvert. 
```{r, message=FALSE, fig.width= 11}
compare_df(
    df_extractor_1 = df.compare1,
    df_extractor_2 = df.compare2,
    output = "html")
```

In grey, values that are consistent between the two data extractors. In green/red, the values that differ.  

Only rows with differences between the two datasets are identified, and you can easily retrieve the row number by looking at the ID in the rowname column. If your datasets have rows in a different order (as in the example above), you can reorder them automatically by indicating to the function the columns that store this information.

```{r, message=FALSE, fig.width= 11}
compare_df(
    df_extractor_1 = df.compare1,
    df_extractor_2 = df.compare2,
    ordering_columns = c("author", "year"),
    output = "html")
```


# Step 3. Effect size computation

## Basic usage

To generate an effect size from a dataset that contains appropriate column names and information, you simply need to:<ul>
<li> pass this dataset to the **convert_df()** function </li>
<li> indicate the effect size measure that should be estimated </li>
</ul>

For this example, we will generate effect sizes from the **df.short** dataset, and we will estimate Hedges' g.

```{r, message=FALSE}
res = convert_df(x = df.short, measure = "g")
```

```{r, eval= FALSE}
summary(res)
```

```{r echo=FALSE}
show_table(summary(res), options = list(  
    scrollX = TRUE,
    dom = c('t'),
    ordering = FALSE,
    rownames = FALSE,
    scrollY = "300px", 
    pageLength = 100,
    columnDefs = list(
                  list(width = '130px',
                       targets = "_all"),
                  list(className = 'dt-center', 
                                     targets = "_all"))))
```

To know more about the information stored in each column, refer to documentation of the summary.metaConvert function, available in the R manual of this package.

## More advanced usage

A tutorial on a more advanced usage is given in the companion paper of this tool: Gosling CJ, Cortese S, Solmi M, et al. metaConvert: an automatic suite for estimation of 11 different effect size measures and flexible conversion across them. *Research Synthesis Methods*. 2025. <a href="https://doi.org/10.1017/rsm.2025.11" target="_blank">doi:10.1017/rsm.2025.11</a>

You can also refer to the documentation of the convert_df function in the R manual of this package, in which all options are described.

## Risk difference and NNT

Version 1.1.0 added the **risk difference** (`measure = "rd"`) as a measure in its own right, and gave the **number needed to treat** (`measure = "nnt"`), available since 1.0.3, a standard error and a confidence interval; before that its uncertainty was always `NA`. When the RD CI crosses zero, the NNT CI is set to NA to preserve the Altman (1998) discontinuous-CI convention - the point estimate and SE remain available.

Only the rows of `df.haza` that hold a complete 2x2 table can yield a risk difference, so the estimable rows are selected before printing.

```{r, message=FALSE}
res_rd <- convert_df(df.haza, measure = "rd")
s_rd <- summary(res_rd)
head(s_rd[!is.na(s_rd$es_crude),
          c("row_id", "es_crude", "se_crude",
            "es_ci_lo_crude", "es_ci_up_crude", "info_used_crude")])
```

## Quality flags and missing-data guidance

`summary()` returns two diagnostic columns by default. The flags column (`flags_crude` and `flags_adjusted` when crude and adjusted estimates are split) reports validation errors (Tier 1: input integrity checks such as negative SDs or inverted CIs) and plausibility warnings (Tier 2: e.g. implausibly large SMD, RD outside `[-1, 1]`, ES/SE outliers across rows). The guidance column (`es_guidance_crude` / `es_guidance_adjusted`) explains, for rows where the effect size could not be estimated, which input columns would have to be added to unlock additional estimators. Both are controlled by the `flags` and `guidance` arguments of `summary()`, which default to `TRUE`; set either to `FALSE` to omit the corresponding column.

```{r, message=FALSE}
res_flags <- convert_df(df.short, measure = "g")
s_flags <- summary(res_flags)
head(s_flags[nzchar(s_flags$flags_crude),
             c("row_id", "es_crude", "info_used_crude", "flags_crude")], 3)
```

The guidance column is only filled for rows without an effect size. In `df.short` such rows carry no usable input at all, so the example below builds a small dataset in which the second comparison reports means and sample sizes but no measure of variability:

```{r, message=FALSE}
df_gap <- data.frame(
  study_id     = c("Study A", "Study B"),
  n_exp        = c(40, 35),    n_nexp       = c(42, 38),
  mean_exp     = c(12.4, 9.8), mean_nexp    = c(10.1, 9.9),
  mean_sd_exp  = c(3.2, NA),   mean_sd_nexp = c(3.4, NA)
)
summary(convert_df(df_gap, measure = "g"))[, c("row_id", "es_crude",
                                               "info_used_crude",
                                               "es_guidance_crude")]
```

## Correction for attenuation due to measurement error

`es_disattenuate()` corrects an observed correlation for unreliability in one or both measures (Spearman, 1904; Hunter & Schmidt, 2004) and propagates the uncertainty into the corrected SE and Fisher's z SE via the delta method. Apply it **after** `summary(convert_df(..., measure = "r"))`:

```{r, message=FALSE}
# Observed correlation with known scale reliabilities
es_disattenuate(r = 0.45, r_se = 0.08,
                reliability_x = 0.82, reliability_y = 0.78)
```

Set `reliability_y = 1.0` if only one side of the correlation needs correcting (e.g. a reliable criterion measure).

## Psychometric utilities: SEM and SDC

Standalone functions compute the **standard error of measurement** and the **smallest detectable change** for meta-analyses of measurement properties. They chain naturally:

```{r}
sem_res <- compute_sem(sd = 10, icc = 0.85, n_sample = 100)
sdc_res <- compute_sdc(sem = sem_res$sem, sem_se = sem_res$sem_se)
sem_res
sdc_res
```

# I do not like R

If you prefer having a graphical user interface (GUI) when performing data analysis, 
we are pleased to introduce you to our web app, which enables you to perform all calculations of this
package using an interactive GUI <a href="https://metaconvert.org/" target="_blank">https://metaconvert.org/</a>

---

